---
title: "Potential Confounders Analysis Pipeline"
output:
  html_document:
    number_sections: true
    toc: true
    toc_float: true
    code_folding: hide
---

Due to IRB restrictions, we cannot share the data needed to demonstrate the code. We can share the R analysis pipeline post data-processing as an example of how to implement our method. The existing code is developed on R version 4.0.2. The relevant package dependences are listed in the next section. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, progress = FALSE)
set.seed(123)

# Relevant packages dependencies
library("survival")
library("ggfortify")
library("glmnet")
library("grf")
library("dplyr")
library("xtable")
library("stringr")
library("ggplot2")
library("MatchIt")
library("cobalt")
library("gridExtra")

# Set variables for analysis
# Directory where data is located
data_dir = "/share/pi/rubin/jiaming/data_confounder"

# Tags to select for cancer type and treatments to analyze
cancer_type = "prostate"
treatA = "surgery"
treatB = "radiation"
tag = "final"
```

### Data Preprocessing

```{r, echo=FALSE}
# Load training and test data with bow matrices. 
data_train = read.csv(sprintf("%s/%s_%s_%s_train.csv", data_dir, cancer_type, treatA, treatB))
data_test = read.csv(sprintf("%s/%s_%s_%s_test.csv", data_dir, cancer_type, treatA, treatB))

# Renumber the treatment column so we have 0/1 encoding
if (treatA == "monitoring") {
    data_train$W[data_train$W == 2] = 0
    data_test$W[data_test$W == 2] = 0
} else if (treatA == "surgery" & treatB == "monitoring") {
    data_train$W[data_train$W == 2] = 1
    data_test$W[data_test$W == 2] = 1
}

# Combine data for train and test and select some columns
data = rbind(data_train, data_test)

# Add character label for treatment
data$W. = data$W
data$W.[data$W. == 0] = treatA
data$W.[data$W. == 1] = treatB

# Select covariates and scale X
covariates = colnames(data)[!colnames(data) %in% c("W", "W.", "Y", "D", "PATIENT_ID")]
X = as.matrix(scale(data[, covariates]))
```

## Plot propensity scores before analysis

We plot the propensity score before analysis to check for overlap and ensure the data is suitable for causal analysis.

```{r}
# Estimate the propensity scores with grf
forest.w = regression_forest(X, data$W, tune.parameters = "all")
W.hat <- predict(forest.w)$predictions
```
```{r}
# Plot the propensity scores with y as density 
prop.orig.plot = PropensityPlot(data$W., W.hat)
ggsave(sprintf("%s_%s_propensity.png", treatA, treatB), width = 7, height = 5, units = "in")
```

## Identifying potential confounders

We identify potential confounders from the data by creating a prediction model for the treatment and outcome respectively. Then, we identify the covariates that are not zero-ed out by both prediction models as the intersection terms, i.e. potential confounders.

```{r}
# Training Lasso model for predicting treatment
treatment_lambda = cv.glmnet(X, data$W, alpha = 1, 
                             family = 'binomial', 
                             nfolds=5)
```
```{r}
# Identify the top predictors of treatment
coefs = coef(treatment_lambda, s = treatment_lambda$lambda.1se)
glmnet.wordimp = as.data.frame(cbind(covariates, coefs[-1]))
names(glmnet.wordimp)[2] <- "coefs"
glmnet.wordimp$coefs = as.numeric(glmnet.wordimp$coefs)

topw.glmnet = glmnet.wordimp[glmnet.wordimp$coefs != 0,]
row.names(topw.glmnet) <- NULL
```
```{r}
treat.coef.plot = CoefficientPlot(topw.glmnet, treatA, treatB, "Treatment")
```


```{r}
# Training a Lasso model for predicting survival
lm.survival = cv.glmnet(X, Surv(data$Y, data$D), 
                        family="cox", type.measure = "C", alpha=1)
```
```{r}
# Find max coefficients from survival model
lm.survival.full = glmnet(X, Surv(data$Y, data$D), 
                          family="cox", alpha=1, type.measure = "C", lambda = lm.survival$lambda.1se)
coxnet.wordimp = as.data.frame(cbind(covariates, as.numeric(lm.survival.full$beta)))
names(coxnet.wordimp)[2] <- "coefs"

coxnet.wordimp$coefs = as.numeric(coxnet.wordimp$coefs)
topw.coxnet = coxnet.wordimp[coxnet.wordimp$coefs != 0,]
rownames(topw.coxnet) <- NULL
```
```{r}
outcome.coef.plot = CoefficientPlot(topw.coxnet, treatA, treatB, "Survival Outcome")
```

We find the intersection terms with the highest likelihood of being potential confounders.

```{r}
# Save coefficient plots
ggsave(file=sprintf("%s_%s_%s_coefs_treatment.png", treatA, treatB, tag), treat.coef.plot, width = 8, height = 5, units = "in")
ggsave(file=sprintf("%s_%s_%s_coefs_outcome.png", treatA, treatB, tag), outcome.coef.plot, width = 8, height = 5, units = "in")
```

## Perform analysis with the combinations of intersection covariates

We utilize the following covariate combinations for comparison:
  - Structured
  - Intersection
  - Intersect+structured
  
We perform the following with different covariate combinations:
  - Multivariate coxph modeling
  - Matching on Propensity Scores
  - Inverse Propensity of Treatment Weighting

For each of the settings above, we plot the 
  - Propensity score with selected covariates
  - Covarite balance plot after weighting
  - The hazard ratio comparison plot
  
```{r}
# Covariate combinations
structured = c('race_white', 'race_api', 'race_black', 'hispanic','pat_age',
               'stage','tumor_grade',
               'grade_unknown',
               'diagnosis_year')
w.intersect = intersect(topw.coxnet$covariates, topw.glmnet$covariates)
intersect.structured = union(w.intersect, structured)
```

```{r}
structured.results = SurvivalAnalysisPipe(structured, data, treatA, treatB, "structured")
intersect.results = SurvivalAnalysisPipe(w.intersect, data, treatA, treatB, "intersect")
intersect.structured.results = SurvivalAnalysisPipe(intersect.structured, data, treatA, treatB, "intersect.structured")
```

# Summarize results and plot data

```{r}
# Create results summary table
results.summary = bind_rows(intersect.structured.results$results, 
                            intersect.stage.results$results, 
                            intersect.results$results, 
                            structured.results$results)
results.summary = results.summary %>%
                      mutate(ID=row_number(),
                            pval=ifelse(treat.p >= .001, sprintf("%.3f", treat.p), "<0.001"),
                            hr.str= sprintf("%s: %.2f (%.2f-%.2f), %s", data_type, HR, HR.ci.lower, HR.ci.upper, pval))
```
```{r}
forestplot = forest.plot(results.summary)
textplot = text.col.plot(results.summary)
g = arrangeGrob(textplot, forestplot, ncol = 2)
ggsave(file=sprintf("%s_%s_%s_summary.png", treatA, treatB, tag), g, width = 10, height = 5, units = "in")
```

Helper Functions for all survival analysis
```{r}
SurvivalAnalysisPipe = function(selected.covariates, data, treatA, treatB, data_type, save_plot=TRUE) {
  df.results = data.frame(
    group = character(),
    data_type = character(),
    model = character(),
    HR = double(),
    HR.ci.lower = double(),
    HR.ci.upper = double(),
    treat.p = double(),
    robust.scr = double(),
    concordance = double(),
    model.p = double(),
    stringsAsFactors = FALSE
  )
  
  # Propensity score of selected covariates
  X.filtered = as.matrix(scale(data[selected.covariates]))
  
  # Plot propensity scores with grf
  forest.w = regression_forest(X.filtered, data$W, tune.parameters = "all")
  propscrs.adjusted <- predict(forest.w)$predictions
  propplot.adjusted = PropensityPlot(data$W., propscrs.adjusted)
  
  # Multivariate coxph model
  sumx = paste(selected.covariates, collapse = " + ")
  surv.formula <- paste("Surv(Y, D) ~ W.", sumx, sep = " + ")
  surv.formula <- as.formula(surv.formula)
  
  coxph = coxph(surv.formula, data = data, robust = TRUE)
  coxph.stats = summary(coxph)
  df.results = rbind(
    df.results,
    gather_coxph_stats(coxph.stats, treatA, treatB, data_type, "multivariate.coxph")
  )
  
  ##### Matching
  # Matching based on selected covariates
  treat.formula <- paste("W", sumx, sep = " ~ ")
  treat.formula <- as.formula(treat.formula)
  m.out <- matchit(treat.formula, data = data, method = "nearest")
  
  matching.balance.plot = love.plot(bal.tab(m.out, s.d.denom = "pooled", disp.means = TRUE, disp.sds = TRUE),
                                    drop.distance = TRUE,
                                    threshold = .2)
  
  # Univariate model in matched sample
  coxph = coxph(as.formula("Surv(Y, D) ~ W."),
                data = match.data(m.out),
                robust = TRUE)
  match.stats = summary(coxph)
  df.results = rbind(
    df.results,
    gather_coxph_stats(match.stats, treatA, treatB, data_type, "matching")
  )
  
  ##### IPTW
  # Inverse Propensity of Treatment Weighting with stabilization
  stab = 0.95
  weights = propscrs.adjusted
  weights = data$W + (1 - data$W) * weights / (1 - weights)
  weights[weights > stab * max(weights)] = stab * max(weights)
  weights[weights < stab * min(weights)] = stab * min(weights)
  data$prop_weights = weights
  
  iptw.balance.plot = love.plot(
    x = treat.formula,
    data = data,
    weights = data$prop_weight,
    method = "weighting",
    drop.distance = TRUE,
    threshold = .2,
    s.d.denom = "pooled"
  )
    
  # Univariate model in IPTW models
  coxph = coxph(
    as.formula("Surv(Y, D) ~ W."),
    data = data,
    robust = TRUE,
    weights = data$prop_weights
  )
  iptw.stats = summary(coxph)
  df.results = rbind(df.results,
                     gather_coxph_stats(iptw.stats, treatA, treatB, data_type, "IPTW"))
  
  if (save_plot) {
    g = arrangeGrob(propplot.adjusted, matching.balance.plot, iptw.balance.plot, ncol = 3)
    ggsave(file=sprintf("%s_%s_%s_%s.png", treatA, treatB, tag, gsub("\\.", "_", data_type)), g, width = 20, height = 5, units = "in")
  }
  
  return (list(results = df.results,
    adj.prop.plot = propplot.adjusted,
    matching.bal.plot = matching.balance.plot,
    iptw.bal.plot = iptw.balance.plot
  ))
}
  
```

Define helper functions for displaying results
```{r helper functions}
PropensityPlot <- function(Treatment, propensity_scores){
  df = data.frame(Treatment = data$W., propensity_scores = propscrs.adjusted)
  ggplot(df, aes(x = propensity_scores, group=Treatment, color=Treatment)) + 
  geom_histogram(fill="white", aes(y=..density..), position="identity", alpha=0.25, bins = 50) + 
  labs(title = "Propensity Scores (P[Treatment|Covariates-Outcomes])",
       x = "Probability of Treatment",
       y = "Density")
}

# Plot the top predictors of treatment and outcome
CoefficientPlot <- function(df, treatA, treatB, outcome) {
  # Set parameters based on model
  if (outcome == "Treatment") {
    yticks = c(min(df$coefs)/2, max(df$coefs)/2)
    ylabels = c(treatA,treatB)
    expandy = c(min(df$coefs)/2, max(df$coefs)/2)
    nudgey = 0.05
  } else {
    yticks = c(min(-0.1, min(df$coefs)/2), max(df$coefs)/2)
    ylabels = c("good prognostic","bad prognostic")
    #expandx = c(0, nrow(df))
    expandy = c(min(-0.1, min(df$coefs)/2), max(df$coefs)/2)
    nudgey = 0.03
  }
  
  # Generate plot
  ggplot(df, aes(x=covariates, y=coefs)) +
  geom_point(aes(colour = factor(df$coefs > 0))) + 
  geom_text(label=df$covariates, nudge_y = nudgey) + 
  geom_hline(yintercept=0, linetype="dashed", color="red") + 
  scale_y_continuous(breaks=yticks,
                   labels=ylabels) + 
  expand_limits(x=c(-3, nrow(df)+3), y=expandy) + 
  theme(#axis.title.x=element_blank(),
        #axis.title.y=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.position="none") + 
  labs(title = sprintf("Coefficients of %s Model: %s-%s", outcome, treatA, treatB),
       y = "Regression Coefficient Indication",
       x = "Covariates")
}

forest.plot <- function(df, title=" ", trt.cols=NULL, xlabel="Hazard Ratio",
                        titles=1, axis=1, shape.leg=TRUE) {
  hr.lims = c(min(results.summary$HR.ci.lower), max(results.summary$HR.ci.upper))
  forest.plt <- df %>%
    ggplot(aes(x=ID, y=HR, color=data_type, shape=model)) +
    geom_hline(aes(yintercept=1), size=.5, linetype=2) +
    geom_point(size=2) +
    geom_linerange(aes(ymin=HR.ci.lower, ymax=HR.ci.upper)) +
    scale_y_continuous(limits=hr.lims, labels=do.call(c, lapply(seq(round(hr.lims[1]*5)/5, hr.lims[2], .2),
                                                                function(i) {
                                                                  ifelse(round(i %% .4, digits=1) %% .4 == 0, i, "")
                                                                })),
                       breaks=seq(round(hr.lims[1]*5)/5, hr.lims[2], .2))
  if (!is.null(trt.cols)) {
    forest.plt <- forest.plt + scale_color_manual(values=trt.cols)
  }
  
  forest.plt <- forest.plt +
    xlab(" ") +
    ylab(ifelse(is.null(xlabel), " ", xlabel)) +
    coord_flip() +
    theme_classic() +
    guides(color=FALSE, shape=guide_legend(reverse=TRUE)) +
    theme(legend.position=c(0.3, 1.05),
          legend.direction = "horizontal",
          legend.key.size = unit(0.2, "cm"),
          axis.text.y=element_blank(), 
          axis.ticks.y=element_blank(),
          axis.title.y=element_blank(), 
          panel.grid.major=element_blank(),
          axis.line.y = element_blank(),
          panel.grid.minor=element_blank(),
          plot.margin=grid::unit(c(2,0,0,0), "mm"),
          legend.box.background=element_rect(color="white", fill = "white"),
          rect=element_rect(fill="transparent")) +
    ggtitle(title)
  if (!shape.leg) {
    forest.plt <- forest.plt + guides(shape=FALSE)
  }
  return(forest.plt)
}

text.col.plot <- function(df, title=" ", expand=TRUE) {
  plt <- ggplot(data=df, aes(y=ID, x=1, label=as.character(hr.str))) +
    geom_text(hjust=1, vjust = 1) +
    ylab(NULL) + xlab(" ") +
    ggtitle(title) +
    scale_x_continuous(limits=c(.95, 1)) +
    coord_cartesian(clip="off") +
    theme(plot.title = element_text(hjust = 0.3, size=5), ## centering title on text
          axis.text.x=element_text(color=alpha('white', 0.0)), ## need text to be printed so it stays aligned with figure but white so it's invisible
          axis.line=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks=element_blank(),
          axis.title.y=element_blank(),legend.position="none",
          panel.background=element_blank(),
          panel.border=element_blank(),
          plot.margin=grid::unit(c(0,0,0,0), "mm"),
          panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),
          plot.background=element_blank(),
          rect=element_rect(fill="transparent"))
  if (expand) {
    plt <- plt +
      scale_y_continuous(expand=c(.065,.065))
  }
  return(plt)
}

gather_coxph_stats = function(coxph.stats, treatA, treatB, data_type, model_name, sig.digits = 3) {
  results = list(group = sprintf("%s-%s", treatA, treatB),
    data_type = data_type,
    model = model_name,
    HR = signif(coxph.stats$conf.int['W.surgery',"exp(coef)"], sig.digits),
    HR.ci.lower = signif(coxph.stats$conf.int['W.surgery',"lower .95"], sig.digits),
    HR.ci.upper = signif(coxph.stats$conf.int['W.surgery',"upper .95"], sig.digits),
    treat.p =  signif(coxph.stats$coefficients['W.surgery','Pr(>|z|)'], sig.digits),
    robust_scr = signif(coxph.stats$robscore[['test']], sig.digits),
    concordance = signif(coxph.stats$concordance[['C']], sig.digits),
    model.p = signif(coxph.stats$robscore[['pvalue']], sig.digits)
  )
  
  return (results)
}
```
